An Efficient Forecasting Approach\\ to Reduce Boundary Effects in Real-Time Time-Frequency Analysis

### 2021-02-22 Paper/BoundEffRed.tex
final version

A. Meynard is the corresponding author (e-mail: adrien.meynard@duke.edu).
}}

@@

Windowing is a common method for performing local analysis. Among many others, STFT~\cite{Flandrin:1999}, continuous wavelet transform (CWT)~\cite{Da1992}, synchrosqueezing transform (SST) \cite{Daubechies11synchrosqueezed}, and reassignment~\cite{Auger13time} (RS) are representations that fall back on the use of an analysis window. Let $x:I\to\RR$ denote the observed signal, where $I$ denotes the finite interval where the signal is measured. Let $g_s:\RR\to\RR$ denote the analysis window, where $s$ is a shape parameter. The support of $g_s$ is localized around the origin and is small with respect to $|I|$. The translation operator is $T_\tau$ defined as:
\[
T_\tau f = f(t-\tau)\ ,\quad \forall\ f:\RR\to\RR\ .
\]
Then, the local analysis of $x$ around the instant $\tau\in I$ rely on the evaluation of the following inner product:
\begin{equation}
V_x(s,\tau) = \langle x, T_\tau g_s \rangle_I \ .
\label{eq:windowing}
\end{equation}
A major shortcoming of this technique occurs when analyzing the signal $x$ {\em near the boundaries} of the interval $I$. Clearly, at these points, half of the information is missing. Consequently, the results of the inner product~\eqref{eq:windowing} are distorted. This phenomenon is usually understood as the \emph{boundary effect}. The result of the SST of a PPG is shown in the lower-left corner of Fig.~\ref{fig:ex.intro} (see section~\ref{ssse:ppg} for a comprehensive description). The distortion resulting from boundary effects is clearly visible on the right side of this representation---this area is highlighted by the red dashed rectangle. Indeed, while in the major left part of the image, clear lines stand out, they become blurred as they approach the right boundary of the image. Therefore, estimations of signal characteristics, like instantaneous frequencies~\cite{Delprat92asymptotic} or amplitudes, from this TF representation appear to be imprecisely determinable (or even likely to fail) in the vicinity of the boundaries. Moreover, this boundary effect would unavoidably limit applying these window-based TF analysis tools for real-time analysis purposes. It is thus desirable to have a solution to eliminating the boundary effects.

@@

Attempts to minimize the boundary effects generally consists in softening the discontinuity on signal edges. Usually, the approaches can be classified into two main classes, choosing a proper analysis window and extending/forecasting the signal. In the first class, judicious choice of analysis window whose support does not interfere with the boundary points can minimize the occurrence of aberrant patterns near the boundaries of the TF plane, for instance, \cite{Chui92wavelets,Depczynski99fast}. Due to the specific relationship between the chosen analysis windows and the TF analysis tool, these techniques do not make it possible to reduce the boundary effects of all TF representations. 
%
Another natural idea, the second class, consists of carrying out a preliminary step of extending, or forecasting, the signal beyond its boundaries. Due to its flexibility, various forecasting schemes have been proposed. For example, there exist simple extension schemes that do not take into account the dynamical behavior of the signal, such as zero-padding, periodic extension, symmetric extension~\cite{Kharitonenko02wavelet,Chen95symmetric}, or polynomial extrapolation~\cite{Williams97discrete}. 
%
There exist extension schemes based on physically relevant dynamical models, such as the Extended Dynamic Mode Decomposition~\cite{Williams15data} (EDMD) and the Gaussian process regression~\cite{Rasmussen06gaussian,Roberts13Gaussian} (GPR).
%
In speech processing, dynamic mode predictors have also been proposed~\cite{Vargas11speech} to forecast signals falling into the so-called \textit{source--filter} model. This gradient-based technique rely on the shadowing approach proposed in~\cite{Grebogi90shadowing}.
%
There also exist extension schemes based on stochastic models, such as the Trigonometric, Box-Cox transformation, ARMA errors, Trend and Seasonal components (TBATS) algorithm~\cite{DeLivera11forecasting}, the dynamic linear models \cite{west2006bayesian}. 
%
In the physically relevant dynamical models, the oscillation or trend, are usually modeled as the mean of a stochastic process, while in the stochastic models, the oscillation or trend are modeled by the covariance structure of the stochastic process. 
It is also possible to consider polynomial regression~\cite{fan1996local} or modeling the mean by splines~\cite{hall2005theory}, kernel functions~\cite{chang2010training}, and wavelets~\cite{marron1998exact}, or nonparametric regression~\cite{fan1996local}, to estimate the mean function before forecasting the signal. Neural network models, such as the long short-term memory model~\cite{vlachas2018data}, is another approach. 
%
The above list is far from exhaustive, and we refer readers with interest to \cite{hyndman2018forecasting} for a friendly monograph on the general forecasting topic. 
%
While the model-based extension scheme gives better-extended signals than those simple extension without dynamical models, they generally have a great computational cost.

@@

The paper is organized in the following way. In section~\ref{se:algo}, we provide an extension method based on a linear dynamic model. We derive the corresponding algorithm for boundary effects reduction. In section~\ref{se:theoretical}, we show that the dynamic model we consider is sufficient to extend signals taking the form of sums of sine waves. An evaluation of the theoretical performance of our algorithm on this class of signals is given in section~\ref{se:theoretical}. In section~\ref{se:results}, we compare our extension method with more sophisticated methods such as EDMD, GPR, or TBATS. We show that our algorithm gives fast results of reasonable quality. Finally, we evaluate the performance of our boundary effects reduction algorithm on biomedical signals, such as respiratory signals, and compare it to the theoretical results.

>>>>

A. Meynard is the corresponding author (e-mail: adrien.meynard@duke.edu).

This paper has supplementary downloadable material available at http://ieeexplore.ieee.org., provided by the author. The material includes a technical report that contains proofs of mathematical statements
and additional illustrations. This material is 1.2~MB in size.
}}

@@

Windowing is a common method for performing local analysis. Among many others, STFT~\cite{Flandrin:1999}, continuous wavelet transform (CWT)~\cite{Da1992}, synchrosqueezing transform (SST) \cite{Daubechies11synchrosqueezed}, and reassignment~\cite{Auger13time} (RS) are representations that fall back on the use of an analysis window. Let $x:I\to\RR$ denote the observed signal, where $I$ denotes the finite interval where the signal is measured. Let $g_s:\RR\to\RR$ denote the analysis window, where $s$ is a shape parameter. The support of $g_s$ is localized around the origin and is small with respect to $|I|$. The translation operator is $T_\tau$ defined as:
\[
T_\tau f = f(t-\tau)\ ,\quad \forall\ f:\RR\to\RR\ .
\]
Then, the local analysis of $x$ around the instant $\tau\in I$ relies on the evaluation of the following inner product:
\begin{equation}
V_x(s,\tau) = \langle x, T_\tau g_s \rangle_I \ .
\label{eq:windowing}
\end{equation}
A major shortcoming of this technique occurs when analyzing the signal $x$ {\em near the boundaries} of the interval $I$. Clearly, at these points, half of the information is missing. Consequently, the results of the inner product~\eqref{eq:windowing} are distorted. This phenomenon is usually understood as the \emph{boundary effect}. The result of the SST of a PPG is shown in the lower-left corner of Fig.~\ref{fig:ex.intro} (see Section~\ref{ssse:ppg} for a comprehensive description). The distortion resulting from boundary effects is clearly visible on the right side of this representation---this area is highlighted by the red dashed rectangle. Indeed, while in the major left part of the image, clear lines stand out, they become blurred as they approach the right boundary of the image. Therefore, estimations of signal characteristics, like instantaneous frequencies~\cite{Delprat92asymptotic} or amplitudes, from this TF representation appear to be imprecisely determinable (or even likely to fail) in the vicinity of the boundaries. Moreover, this boundary effect would unavoidably limit applying these window-based TF analysis tools for real-time analysis purposes. It is thus desirable to have a solution to eliminating the boundary effects.

@@

Attempts to minimize the boundary effects generally consists in softening the discontinuity on signal edges. Usually, the approaches can be classified into two main classes, choosing a proper analysis window and extending/forecasting the signal. In the first class, judicious choice of analysis window whose support does not interfere with the boundary points can minimize the occurrence of aberrant patterns near the boundaries of the TF plane, for instance, \cite{Chui92wavelets,Depczynski99fast}. Due to the specific relationship between the chosen analysis windows and the TF analysis tool, these techniques do not make it possible to reduce the boundary effects of all TF representations. 
%
Another natural idea, the second class, consists of carrying out a preliminary step of extending, or forecasting, the signal beyond its boundaries. Due to its flexibility, various forecasting schemes have been proposed. For example, there exist simple extension schemes that do not take into account the dynamical behavior of the signal, such as zero-padding, periodic extension, symmetric extension~\cite{Kharitonenko02wavelet,Chen95symmetric}, or polynomial extrapolation~\cite{Williams97discrete}. 
%
There exist extension schemes based on physically relevant dynamical models, such as the Extended Dynamic Mode Decomposition~\cite{Williams15data} (EDMD) and the Gaussian process regression~\cite{Rasmussen06gaussian,Roberts13Gaussian} (GPR).
%
In speech processing, dynamic mode predictors have also been proposed~\cite{Vargas11speech} to forecast signals falling into the so-called \textit{source--filter} model. This gradient-based technique relies on the shadowing approach proposed in~\cite{Grebogi90shadowing}.
%
There also exist extension schemes based on stochastic models, such as the Trigonometric, Box-Cox transformation, ARMA errors, Trend and Seasonal components (TBATS) algorithm~\cite{DeLivera11forecasting}, the dynamic linear models \cite{west2006bayesian}. 
%
In the physically relevant dynamical models, the oscillation or trend, are usually modeled as the mean of a stochastic process, while in the stochastic models, the oscillation or trend are modeled by the covariance structure of the stochastic process. 
It is also possible to consider polynomial regression~\cite{fan1996local} or modeling the mean by splines~\cite{hall2005theory}, kernel functions~\cite{chang2010training}, and wavelets~\cite{marron1998exact}, or nonparametric regression~\cite{fan1996local}, to estimate the mean function before forecasting the signal. Neural network models, such as the long short-term memory model~\cite{vlachas2018data}, is another approach. 
%
The above list is far from exhaustive, and we refer readers with interest to \cite{hyndman2018forecasting} for a friendly monograph on the general forecasting topic. 
%
While the model-based extension scheme gives better-extended signals than those simple extension without dynamical models, they generally have a great computational cost.

@@

The paper is organized in the following way. In section~\ref{se:algo}, we provide an extension method based on a linear dynamic model. We derive the corresponding algorithm for boundary effects reduction. In Section~\ref{se:theoretical}, we show that the dynamic model we consider is sufficient to extend signals taking the form of sums of sine waves. An evaluation of the theoretical performance of our algorithm on this class of signals is given in Section~\ref{se:theoretical}. In Section~\ref{se:results}, we compare our extension method with more sophisticated methods such as EDMD, GPR, or TBATS. We show that our algorithm gives fast results of reasonable quality. Finally, we evaluate the performance of our boundary effects reduction algorithm on biomedical signals, such as respiratory signals, and compare it to the theoretical results.

### 2021-02-22 Paper/algorithm.tex
colors
final version

\paragraph{Dynamical Model} 
Establishing a dynamical model means determining the relation linking $\bY$ to $\bX$; that is, finding a function $f$ so that
\begin{equation}
\bY = f(\bX)\ .
\label{eq:generic.model}
\end{equation}
In a general framework, forecasting means estimating the function $f$ from the observed values taken by the signal, in order to predict its future values. For instance, the dynamic mode decomposition~\cite{Schmid10dynamic,Williams15data} or other more complicated models \cite{Roberts13Gaussian,DeLivera11forecasting,west2006bayesian,vlachas2018data}, allow this by setting additional constraints on the behavior of $f$. We will see, in section~\ref{se:theoretical}, that considering such a complex dynamic model is not necessary for the study of the oscillatory signals of interest to us. That is why we consider here a naive dynamical model, assuming that we have the following relation:
\begin{equation}
\bY = \bA\bX\ ,
\label{eq:dyn.model}
\end{equation}
where $\bA\in\RR^{M\times M}$. In other words, we adopt a classical strategy in the study of dynamical systems, the {\em linearization of a nonlinear system}, when the system is sufficiently regular. Notice that this linearized dynamical model can be written equivalently according to the subsignals $\bx_k$, as:
\begin{equation}
\bx_{k+1} = \bA\bx_{k},\ ,\forall k\in\{0,\dots,K-1\}\ .
\end{equation}
%
%{\color{blue}
%From the Takens' embedding perspective~\cite{Takens81detecting}, $\bX$ can be viewed as a lag map. Under the manifold assumption of the intrinsic phase space that hosts the dynamics, $f$ describes the dynamics supported on the manifold space, and $\bA$ approximates the dynamics on the manifold.
%}

{\color{red}
\paragraph{Forecasting}
Determining the forecasting amounts to estimating the unknown matrix $\bA$. Indeed, let $\tilde\bA$ denotes the estimate of $\bA$. 
%
We then obtain the forecasting of $\bx_{k+\ell}$ by recursively applying the linear relation~\eqref{eq:dyn.model}:
\begin{equation*}
\tilde\bx_{K+\ell} = \underbrace{\tilde\bA\tilde\bA\cdots\tilde\bA}_{\scriptstyle\ell\mbox{ \footnotesize {times}}}\bx_{K} = \tilde\bA^\ell\bx_{K}\ .
\end{equation*}  
Let $\be_M$ denote the unit vector of length $M$ given by $\be_M = \begin{pmatrix} 0 & \cdots & 0 & 1\end{pmatrix}^T$. Consequently, given definition~\eqref{eq:xk}, the forecasting of the signal at time $\frac{N-1+\ell}{\fs}$ is provided by
\begin{equation}
\label{eq:prediction}
\tilde \bx [N-1+\ell] = \be_M^T\tilde\bx_{K+\ell} = \balpha^{(\ell)}\bx_{K}\ ,
\end{equation}  
where $\balpha^{(\ell)}$ denotes the last row of $\tilde\bA^\ell$, \ie~$\balpha^{(\ell)}=\be_M^T\tilde\bA^\ell$.
}

\paragraph{Model Estimation} To estimate the matrix $\bA$, we consider the simple but numerically efficient least square estimator. That is, we solve the following problem:
\begin{equation}
\label{eq:ls.pb}
\tilde\bA = \arg\min_{\balpha} \cL(\bA)\ ,
\end{equation}
where the loss function $\cL$ is given by:
\[
\cL(\bA) = \|\bY-\bA\bX\|^2 = \sum_{k=0}^{K-1} \|\bx_{k+1}-\bA\bx_k\|^2.
\]
Therefore, solving the problem~\eqref{eq:ls.pb}, \ie~$\nabla \cL(\tilde\bA)=\bzero$, gives the following estimate $\tilde\bA$ of the dynamical model matrix $\bA$:
\begin{align}
\tilde\bA &= \bY\bX^T(\bX\bX^T)^{-1}\ .
\label{eq:lse}
\end{align}

>>>>

\paragraph{Dynamical Model} 
Establishing a dynamical model means determining the relation linking $\bY$ to $\bX$; that is, finding a function $f$ so that
\begin{equation}
\bY = f(\bX)\ .
\label{eq:generic.model}
\end{equation}
In a general framework, forecasting means estimating the function $f$ from the observed values taken by the signal, in order to predict its future values. For instance, the dynamic mode decomposition~\cite{Schmid10dynamic,Williams15data} or other more complicated models \cite{Roberts13Gaussian,DeLivera11forecasting,west2006bayesian,vlachas2018data}, allow this by setting additional constraints on the behavior of $f$. We will see, in Section~\ref{se:theoretical}, that considering such a complex dynamic model is not necessary for the study of the oscillatory signals of interest to us. That is why we consider here a naive dynamical model, assuming that we have the following relation:
\begin{equation}
\bY = \bA\bX\ ,
\label{eq:dyn.model}
\end{equation}
where $\bA\in\RR^{M\times M}$. In other words, we adopt a classical strategy in the study of dynamical systems, the {\em linearization of a nonlinear system}, when the system is sufficiently regular. Notice that this linearized dynamical model can be written equivalently according to the subsignals $\bx_k$, as:
\begin{equation}
\bx_{k+1} = \bA\bx_{k},\ ,\forall k\in\{0,\dots,K-1\}\ .
\end{equation}
%
%{\color{blue}
%From the Takens' embedding perspective~\cite{Takens81detecting}, $\bX$ can be viewed as a lag map. Under the manifold assumption of the intrinsic phase space that hosts the dynamics, $f$ describes the dynamics supported on the manifold space, and $\bA$ approximates the dynamics on the manifold.
%}

\paragraph{Forecasting}
Determining the forecasting amounts to estimating the unknown matrix $\bA$. Indeed, let $\tilde\bA$ denotes the estimate of $\bA$. 
%
We then obtain the forecasting of $\bx_{k+\ell}$ by recursively applying the linear relation~\eqref{eq:dyn.model}:
\begin{equation*}
\tilde\bx_{K+\ell} = \underbrace{\tilde\bA\tilde\bA\cdots\tilde\bA}_{\scriptstyle\ell\mbox{ \footnotesize {times}}}\bx_{K} = \tilde\bA^\ell\bx_{K}\ .
\end{equation*}  
Let $\be_M$ denote the unit vector of length $M$ given by $\be_M = \begin{pmatrix} 0 & \cdots & 0 & 1\end{pmatrix}^T$. Consequently, given definition~\eqref{eq:xk}, the forecasting of the signal at time $\frac{N-1+\ell}{\fs}$ is provided by
\begin{equation}
\label{eq:prediction}
\tilde \bx [N-1+\ell] = \be_M^T\tilde\bx_{K+\ell} = \balpha^{(\ell)}\bx_{K}\ ,
\end{equation}  
where $\balpha^{(\ell)}$ denotes the last row of $\tilde\bA^\ell$, \ie, $\balpha^{(\ell)}=\be_M^T\tilde\bA^\ell$.

\paragraph{Model Estimation} To estimate the matrix $\bA$, we consider the simple but numerically efficient least square estimator. That is, we solve the following problem:
\begin{equation}
\label{eq:ls.pb}
\tilde\bA = \arg\min_{\balpha} \cL(\bA)\ ,
\end{equation}
where the loss function $\cL$ is given by:
\[
\cL(\bA) = \|\bY-\bA\bX\|^2 = \sum_{k=0}^{K-1} \|\bx_{k+1}-\bA\bx_k\|^2.
\]
Therefore, solving the problem~\eqref{eq:ls.pb}, \ie, $\nabla \cL(\tilde\bA)=\bzero$, gives the following estimate $\tilde\bA$ of the dynamical model matrix $\bA$:
\begin{align}
\tilde\bA &= \bY\bX^T(\bX\bX^T)^{-1}\ .
\label{eq:lse}
\end{align}

### 2021-02-22 Paper/results.tex
final version

\subsection{Simulated Signals}
We first evaluate the quality of the forecasting step and compare it to the theoretical results provided by Theorem~\ref{th:error}. The level of the forecasting error depends on at least two parameters:
\begin{itemize}
\item The noise variance $\sigma^2$.
\item The size of the training dataset $K$. 
\end{itemize}
In subsections~\ref{ssse:res.sine} and~\ref{ssse:res.ahm}, we study the influence of these parameters. A comparison with the theoretical results of section~\ref{se:theoretical} is also available.

@@

\paragraph{Influence of the subsignals length $M$}
To evaluate the forecasting quality, {\sf SigExt} is applied to $1000$ realizations of the above-described signal. We forecast signal extensions of $0.1$~s, \ie~$L=700$. Table~\ref{tab:mse.sine} shows the resulting mean and standard deviation (SD) of the experimental MSE, according to three different values of the subsignals length $M$. Note that choosing a too small $M$ does not provide enough information to forecast the signal satisfactorily. Furthermore, a too large $M$ makes the algorithm insensitive to local nonstationarities produced by frequency variations. The user must then make a compromise on the choice of $M$ in order to optimize the performance of the algorithm. A choice of $M$ such that the subsignals contain about a dozen oscillations is generally appropriate.

\paragraph{Comparison of the extensions methods}
The same set of experiments is run on the extension methods outlined in section~\ref{sse:methods}. The results, displayed in Table~\ref{tab:mse.sine}, show that the naive extension we propose gives satisfying results. Indeed, when the subsignal length $M$ is optimally chosen, {\sf SigExt} performance is of the same order as that of ore sophisticated methods, like GPR or TBATS. Besides, even though these methods could be slightly more robust to perturbations, they are substantially limited by the computing time they require, which prevents them from being used to exploit real-time data. Thus, {\sf SigExt} is the extension method that optimizes the trade-off between forecasting quality and computing time.
%
Besides, the last column of Table~\ref{tab:mse.sine} gives the performance index $D$ of the boundary-free STFT associated with each extension method. This illustrates the ability of our algorithm to reduce boundary effects to a level comparable to what is possible with the GPR or TBATS extensions. The performance of {\sf BoundEffRed} on other TF representations is more thoroughly analyzed on real-life signals in the following.

@@

From that long signal, we build a dataset of $379$ non\-overlapping signals of 60 seconds, \ie~$N=6000$. For each segment, we apply the forecasting methods introduced in section~\ref{ssse:res.ahm}, including the {\sf SigExt} method detailed in Algorithm~\ref{alg:extension}. We forecast $7$-second-long extensions on each segment of the signal, corresponding to $L =700$. Thus, in order to catch the dynamical behavior, the size of the training signal $M$ is chosen so that $M=\lfloor 1.5L\rfloor$. As a result of section~\ref{ssse:res.sine}, we take: $K=\lfloor2.5M\rfloor$. The extensions obtained on one of these subsignals are shown in Fig.~\ref{fig:tho}. The resulting MSEs of different methods are given in Table~\ref{tab:THO}. Note that the results are given in ``mean $\pm$ standard deviation'' format. The TBATS extension method is not implemented, as its excessive computing time makes it impossible to implement it on $6000$ segments within a reasonable period of time.
%
Note that the MSE of {\sf SigExt} is on average higher than the MSE of EDMD or the symmetric extension, with a huge standard deviation. This variability is caused by the presence of few segments contaminated by artifacts so that it is unpredictable via a too simple dynamical model like~\eqref{eq:dyn.model}. The left of Fig.~\ref{fig:THO.failure} illustrates one of these outliers, where {\sf SigExt} fails to catch the fast varying dynamic of the instantaneous amplitude to satisfactorily forecast the signal. The EDMD and symmetric extensions are more robust to those situations, as shown in this example, and in Table~\ref{tab:THO}. Nevertheless, {\sf SigExt} provides a sufficiently relevant extension to give TF representations sparingly affected by boundary effects. On the right of Fig.~\ref{fig:THO.failure}, we display a comparison between the right boundary of SST of the same segment of signal (top-right), and its boundary-free SST obtained after the {\sf SigExt} forecasting (bottom-right). The extension of the instantaneous frequency visible on the right side of the image illustrates the reduction of boundary effects produced despite an inaccurate signal forecasting.

@@

\subsection{Real-Time Implementation}
We simulate the real-time implementation of the SST from {\sf BoundEffRed} applied to the PPG described in section~\ref{ssse:ppg}, subsampled by a factor of $2$. The test is performed on a 2-Core Intel Core i5 CPU running at $1.7$~GHz and $7.7$~GB of RAM.

For this signal, sampled at $\fs=65.5$~Hz, the suitable window is $8$ seconds long. Therefore, the extension must be over $4$ seconds, \ie~$L=250$. The forecasting step in each iteration of {\sf BoundEffRed} then takes no more than $t_{\mathrm{forecast}}=46$~ms. Denote by $H\geq 1$ the hop size in samples, that is the number of samples between successive columns of the SST. Besides, since the support of the analysis window covers $2L$ samples, the update of the SST requires the calculation of $\lceil 2L/H \rceil$ new columns of the SST. In this example, the frequency dimension of the SST being $512$, the computational time of one column is $t_{\mathrm{SST}}=2.08$~ms, on average. A general rule to determine the acceptable values of $H$ for real-time implementation is verifying that the computational time to update the boundary-free TF representation is smaller than the lag between $H$ samples; that is,
\[
t_{\mathrm{forecast}} + \left\lceil \frac{L}H \right\rceil t_{\mathrm{SST}} < \frac{H}\fs\ . 
\]
In our example, taking $H\geq 8$~samples is sufficient to ensure the feasibility of real-time implementation. This value is reasonable because it allows a maximum overlap of $98.4\%$ of the window length. SST variations are therefore perceived gradually. Combined with the low latency of the algorithm, this allows for smooth monitoring of the real-time SST. A video clip illustrating this is available on GitHub, at \url{https://github.com/AdMeynard/BoundaryEffectsReduction/tree/master/Animations}.

>>>>

\subsection{Simulated Signals}
We first evaluate the quality of the forecasting step and compare it to the theoretical results provided by Theorem~\ref{th:error}. The level of the forecasting error depends on at least two parameters:
\begin{itemize}
\item The noise variance $\sigma^2$.
\item The size of the training dataset $K$. 
\end{itemize}
In Sections~\ref{ssse:res.sine} and~\ref{ssse:res.ahm}, we study the influence of these parameters. A comparison with the theoretical results of Section~\ref{se:theoretical} is also available.

@@

\paragraph{Influence of the subsignals length $M$}
To evaluate the forecasting quality, {\sf SigExt} is applied to $1000$ realizations of the above-described signal. We forecast signal extensions of $0.1$~s, \ie, $L=700$. Table~\ref{tab:mse.sine} shows the resulting mean and standard deviation (SD) of the experimental MSE, according to three different values of the subsignals length $M$. Note that choosing too small an $M$ does not provide enough information to forecast the signal satisfactorily. Furthermore, too large an $M$ makes the algorithm insensitive to local nonstationarities produced by frequency variations. The user must then make a compromise on the choice of $M$ in order to optimize the performance of the algorithm. A choice of $M$ such that the subsignals contain about a dozen oscillations is generally appropriate.

\paragraph{Comparison of the extensions methods}
The same set of experiments is run on the extension methods outlined in Section~\ref{sse:methods}. The results, displayed in Table~\ref{tab:mse.sine}, show that the naive extension we propose gives satisfying results. Indeed, when the subsignal length $M$ is optimally chosen, {\sf SigExt} performance is of the same order as that of ore sophisticated methods, like GPR or TBATS. Besides, even though these methods could be slightly more robust to perturbations, they are substantially limited by the computing time they require, which prevents them from being used to exploit real-time data. Thus, {\sf SigExt} is the extension method that optimizes the trade-off between forecasting quality and computing time.
%
Besides, the last column of Table~\ref{tab:mse.sine} gives the performance index $D$ of the boundary-free STFT associated with each extension method. This illustrates the ability of our algorithm to reduce boundary effects to a level comparable with what is possible with the GPR or TBATS extensions. The performance of {\sf BoundEffRed} on other TF representations is more thoroughly analyzed on real-life signals in the following.

@@

From that long signal, we build a dataset of $379$ non\-overlapping signals of 60 seconds, \ie, $N=6000$. For each segment, we apply the forecasting methods introduced in Section~\ref{ssse:res.ahm}, including the {\sf SigExt} method detailed in Algorithm~\ref{alg:extension}. We forecast $7$-second-long extensions on each segment of the signal, corresponding to $L =700$. Thus, in order to catch the dynamical behavior, the size of the training signal $M$ is chosen so that $M=\lfloor 1.5L\rfloor$. As a result of Section~\ref{ssse:res.sine}, we take: $K=\lfloor2.5M\rfloor$. The extensions obtained on one of these subsignals are shown in Fig.~\ref{fig:tho}. The resulting MSEs of different methods are given in Table~\ref{tab:THO}. Note that the results are given in ``mean $\pm$ standard deviation'' format. The TBATS extension method is not implemented, as its excessive computing time makes it impossible to implement it on $6000$ segments within a reasonable period of time.
%
Note that the MSE of {\sf SigExt} is on average higher than the MSE of EDMD or the symmetric extension, with a huge standard deviation. This variability is caused by the presence of few segments contaminated by artifacts so that it is unpredictable via a too simple dynamical model like~\eqref{eq:dyn.model}. The left of Fig.~\ref{fig:THO.failure} illustrates one of these outliers, where {\sf SigExt} fails to catch the fast varying dynamic of the instantaneous amplitude to satisfactorily forecast the signal. The EDMD and symmetric extensions are more robust to those situations, as shown in this example, and in Table~\ref{tab:THO}. Nevertheless, {\sf SigExt} provides a sufficiently relevant extension to give TF representations sparingly affected by boundary effects. On the right of Fig.~\ref{fig:THO.failure}, we display a comparison between the right boundary of SST of the same segment of signal (top-right), and its boundary-free SST obtained after the {\sf SigExt} forecasting (bottom-right). The extension of the instantaneous frequency visible on the right side of the image illustrates the reduction of boundary effects produced despite an inaccurate signal forecasting.

@@

\subsection{Real-Time Implementation}
We simulate the real-time implementation of the SST from {\sf BoundEffRed} applied to the PPG described in Section~\ref{ssse:ppg}, subsampled by a factor of $2$. The test is performed on a 2-Core Intel Core i5 CPU running at $1.7$~GHz and $7.7$~GB of RAM.

For this signal, sampled at $\fs=65.5$~Hz, the suitable window is $8$ seconds long. Therefore, the extension must be over $4$ seconds, \ie, $L=250$. The forecasting step in each iteration of {\sf BoundEffRed} then takes no more than $t_{\mathrm{forecast}}=46$~ms. Denote by $H\geq 1$ the hop size in samples, that is the number of samples between successive columns of the SST. Besides, since the support of the analysis window covers $2L$ samples, the update of the SST requires the calculation of $\lceil 2L/H \rceil$ new columns of the SST. In this example, the frequency dimension of the SST being $512$, the computational time of one column is $t_{\mathrm{SST}}=2.08$~ms, on average. A general rule to determine the acceptable values of $H$ for real-time implementation is verifying that the computational time to update the boundary-free TF representation is smaller than the lag between $H$ samples; that is,
\[
t_{\mathrm{forecast}} + \left\lceil \frac{L}H \right\rceil t_{\mathrm{SST}} < \frac{H}\fs\ . 
\]
In our example, taking $H\geq 8$~samples is sufficient to ensure the feasibility of real-time implementation. This value is reasonable because it allows a maximum overlap of $98.4\%$ of the window length. SST variations are therefore perceived gradually. Combined with the low latency of the algorithm, this allows for smooth monitoring of the real-time SST. A video clip illustrating this is available on GitHub, at \url{https://github.com/AdMeynard/BoundaryEffectsReduction/tree/master/Animations}.
